In \autoref{ch:implementation} we evolved the synthesis judgment of \mlsyn{} into an efficient synthesis procedure.
This procedure went through several iterations in the chapter, so we summarize the final product here.

The synthesis procedure that we implement directly in our prototype program synthesis tool, \myth{} (\autoref{ch:evaluating-myth}), consists of:
\begin{itemize}
  \item A \emph{refinement tree generation algorithm} (\autoref{fig:app-implementation-refinement-tree}),
  \item A \emph{relevant term generation algorithm} (\autoref{fig:app-implementation-genE} and \autoref{fig:app-implementation-genI}), and
  \item An overall synthesis procedure that, given a synthesis specification, synthesizes the first program it finds that satisfies that specification.
\end{itemize}

Given a signature $Σ$, context $Γ$, goal type $τ$, and examples $Χ$, our synthesis procedure operates as follows:
\begin{enumerate}
  \item Let our search parameter initial values be $s = 1$ (the maximum match scrutinee size), $m = 0$ (the maximum match depth), and $k = 13$ (the maximum $E$-term size).
  \item Starting with the initial synthesis $I$-refinement phase of the synthesis plan described below:
    \begin{enumerate}
      \item Execute the next phase of $I$-refinement.
      \item Perform $E$-guessing for 0.25 seconds (using $\mguess$ defined in \autoref{fig:app-implementation-refinement-tree}) up to size $k$ at all elligible nodes in the current refinement tree.
        A refinement tree node is elligible for $E$-guessing if we have not yet found a satisfying expression at that node and its corresponding goal type is a base type $T$.
      \item Propogate newly found satisfying programs upwards in the refinement tree.
        If we construct an overall, satisfying program to the original synthesis problem then terminate, returning that program.
        Otherwise, move onto the next phase of $I$-refinement repeat.
    \end{enumerate}
\end{enumerate}
The phases of $I$-refinement in our synthesis plan are:
\begin{enumerate}
  \item Create an initial refinement tree with the search parameters $s$ and $m$.
  \item Increase $m$ by one ($m = 1$) and extend the refinement tree accordingly.
  \item Increase $m$ by one ($m = 2$) and extend the refinement tree accordingly.
  \item Increase $s$ by five ($s = 6$) and extend the refinement tree accordingly.
  \item Increase $m$ by one ($m = 3$).
\end{enumerate}
If we are unable to find a satisfying program and run out of synthesis steps, we return an error stating that the procedure could not find a satisfying program.

\input{figures/app-implementation}
